### 2024-04-02 paper/paper.qmd
Added introduction.
Added the majority of the introduction that can be added up to this point in the writing process. Also added some references.

---
title: "Widening income gaps and the shrinking of the middle class"
subtitle: "An analysis of the effects of various economic policies in the United States on the standards of living of middle and lower income households."
author: 
  - Abbass Sleiman
thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US-Middle-Class](https://github.com/AbbassSleiman/US-Middle-Class)."
date: today
date-format: long
format: 
  pdf: 
    include-in-header: 
      text: |
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
    documentclass: article
    geometry: margin = 1.2in
    abstract: ""
    number-sections: true
output: 
  bookdown::pdf_document2
toc: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
```

# Introduction

The last 5 decades in the United States saw a clear contraction in the portion of adults residing in middle-class households - a steady decrease from 61% in 1971 to 50% in 2021 [@citepewmiddle]. Concerns regarding affordability of housing has increased drastically in recent years, as evidenced by the fact that 49% of Americans in 2021 found that the availability of affordable housing was a significant challenge [@citepewhouseaffordability]. Between 2017 and 2022, rents rose 18% and the average cost of all goods and services increased 16% [@citepewhouseaffordability].

Given that the middle class continues to comprise the majority of US adults, it is imperative to consider how the standards of living of these individuals, as well as those of lower-income households, have been impacted by these rising costs. As they are the most affected demographic by these economic shifts, understanding the implications on their daily lives, financial stability, and overall well-being is crucial for crafting effective policies and interventions to address these challenges.

In order to best understand the data and literature at hand, it is important to abide by a concrete definition of each of the three primary income brackets, lower-income, middle-income, and upper-income, to ensure consistency throughout the paper, as well as across the various literature cited. Following the definition provided by the Pew Research Center, this paper defines a middle-income adult in a given year as having an annual household income between two-thirds and twice the national median income in the year before, after adjusting for household size [@citepewmiddle]. Consequently, lower-income adults and upper-income adults are defined as those with less than two-thirds and over twice the median income respectively. It is important to note that we adjust for household size to account for the fact that smaller households require a lesser income to support similar lifestyles as larger households.

This paper utilizes data from [...] and seeks to examine how the relative proportions of each income class have changed over the past [...] years, and subsequently evaluate how various economic policy changes - interest rates, taxation, immigration, [...] - have influenced these proportions over time whilst offering a comprehensive analysis of how and why these policies may have caused the changes observed. Moreover, this paper aims to uncover how various individual-level characteristics including race, age, education, and [others], intersect with these policies to shape income distribution dynamics in order to address potential disparities and better explore how best to pursue the goal of economic equity across individuals of all
backgrounds.

This paper finds evidence that [...].

The remainder of this paper is structured as follows. [...].

# Data {#sec-data}

# Model

## Model set-up

### Model justification

# Results

# Discussion

## First discussion point {#sec-first-point}

## Second discussion point

## Third discussion point

## Weaknesses and next steps

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

\newpage

# References

### 2024-04-02 paper/paper.qmd
Updated subtitle

@@
 title: "Widening income gaps and the shrinking of the middle class"
-subtitle: "An analysis of the effects of various economic policies in the United States on the standards of living of middle and lower income households."
+subtitle: "An analysis of the impact of economic policies on the living standards of middle and lower-income households amidst rising costs and affordability challenges."
 author: 

### 2024-04-02 paper/paper.qmd
Added placeholder abstract.

@@
     geometry: margin = 1.2in
-    abstract: ""
+    abstract: "Over the past five decades, the United States has witnessed a decline in the percentage of adults residing in middle-class households, alongside escalating concerns about housing affordability.
+This study aims to assess the impact of rising costs on the living standards of middle and lower-income households, recognizing them as the most affected demographic by these economic shifts.
+Utilizing data from [...], we analyze changes in income distribution and evaluate the influence of economic policy changes, including interest rates and taxation, on the standards of living of these individuals.
+Additionally, we explore how individual-level characteristics intersect with these policies to address how best to pursue the goal of economic equity across individuals of all backgrounds.
+We find that [...]."
     number-sections: true

### 2024-04-15 paper/paper.qmd
Changed paper idea and updated introduction

@@
 ---
-title: "Widening income gaps and the shrinking of the middle class"
-subtitle: "An analysis of the impact of economic policies on the living standards of middle and lower-income households amidst rising costs and affordability challenges."
+title: "Behind Bars and Beyond: Investigating Socioeconomic Drivers of Incarceration Rates in the United States"
+subtitle: "A comparative analysis of global trends and America's mass incarceration crisis."
 author: 
@@
 
-The last 5 decades in the United States saw a clear contraction in the portion of adults residing in middle-class households - a steady decrease from 61% in 1971 to 50% in 2021 [@citepewmiddle].
-Concerns regarding affordability of housing has increased drastically in recent years, as evidenced by the fact that 49% of Americans in 2021 found that the availability of affordable housing was a significant challenge [@citepewhouseaffordability].
-Between 2017 and 2022, rents rose 18% and the average cost of all goods and services increased 16% [@citepewhouseaffordability].
+In recent decades, the United States has grappled with a consistent rise in its prison population, becoming infamous for having one of the highest incarceration rates in the world.
+Leading the world with nearly 2 million individuals in prisons and jails in 2022, the US is faced with what, at first glance, appears to be a "uniquely American problem" [@citeUSincarceration].
+This phenomenon has sparked widespread concern and debate over how best to address this issue.
+Central to these discussions is the question of what underlying factors have contributed to this circumstance, in turn necessitating an exploration of the socioeconomic drivers of incarceration rates in the United States.
 
-Given that the middle class continues to comprise the majority of US adults, it is imperative to consider how the standards of living of these individuals, as well as those of lower-income households, have been impacted by these rising costs.
-As they are the most affected demographic by these economic shifts, understanding the implications on their daily lives, financial stability, and overall well-being is crucial for crafting effective policies and interventions to address these challenges.
+Incarceration refers to the state of being confined in prison as punishment for criminal offences [@citeIncarcerationRate], and in examining the United States' prison dilemma, it becomes clear that the crux of the problem is not easily reducible to simple explanations.
+Instead, it is a complex issue influenced by a multitude of factors including cultural dynamics and socioeconomic conditions.
+Thus, in an attempt to unravel the intricacies of these factors, this paper seeks to conduct a comparative analysis of global trends in education, unemployment, crime, and poverty rates as independent factors influencing a nation's incarceration.
+However, recognizing that the total number of incarcerations is heavily dependent on population size, this paper focuses on nations' incarceration rates as the estimand, defined to be the number of incarcerated individuals per 100,000 in a given population [@citeIncarcerationRate], which serves as a better metric for measuring the extent of imprisonment within a society.
+By exploring the impact of these factors on incarceration rates, we aim to better understand the primary forces behind the United States' high incarceration rate and overall number of individuals incarcerated.
+This understanding will help inform policy-making decisions aimed at addressing the issue effectively.
 
-In order to best understand the data and literature at hand, it is important to abide by a concrete definition of each of the three primary income brackets, lower-income, middle-income, and upper-income, to ensure consistency throughout the paper, as well as across the various literature cited.
-Following the definition provided by the Pew Research Center, this paper defines a middle-income adult in a given year as having an annual household income between two-thirds and twice the national median income in the year before, after adjusting for household size [@citepewmiddle].
-Consequently, lower-income adults and upper-income adults are defined as those with less than two-thirds and over twice the median income respectively.
-It is important to note that we adjust for household size to account for the fact that smaller households require a lesser income to support similar lifestyles as larger households.
+This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Crime Rate by Country 2024".
+By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that [...]
 
-This paper utilizes data from [...] and seeks to examine how the relative proportions of each income class have changed over the past [...] years, and subsequently evaluate how various economic policy changes - interest rates, taxation, immigration, [...] - have influenced these proportions over time whilst offering a comprehensive analysis of how and why these policies may have caused the changes observed.
-Moreover, this paper aims to uncover how various individual-level characteristics including race, age, education, and [others], intersect with these policies to shape income distribution dynamics in order to address potential disparities and better explore how best to pursue the goal of economic equity across individuals of all
-backgrounds.
-
-This paper finds evidence that [...].
-
-The remainder of this paper is structured as follows. [...].
+The remainder of this paper is structured as follows.
+@sec-data discusses the raw data, cleaning process, variables of interest, and offers visual representations of the data through tabular and graphical representations of the observations.
+@sec-model introduces and justifies the Bayesian linear regression model used in the analysis of the data in predicting the relationship between the aforementioned independent variables and a nation's incarceration rate.
+@sec-results deals with analyzing the trends and correlations showcased by the model in more detail.
+@sec-discussion discusses the real-world implications of the results uncovered in the prior sections, and finally @sec-limitations discusses the limitations and weaknesses of the analysis conducted, as well as the next steps that could be taken to improve the overall reliability of the paper.
 
@@
 
-# Model
+# Model {#sec-model}
 
@@
 
-# Results
+# Results {#sec-results}
 
-# Discussion
+# Discussion {#sec-discussion}
 
@@
 
-## Weaknesses and next steps
+# Weaknesses and next steps {#sec-limitations}
 

### 2024-04-15 scripts/01-download_data.R
Added instructions on how to obtain data in download data script

@@ -1,26 +1,19 @@
 #### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Details how to acquire the raw datasets used in this paper.
+# Author: Abbass Sleiman
+# Date: 15 April 2024
+# Contact: abbass.sleiman@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-# [...UPDATE THIS...]
+# Pre-requisites: N/A
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
-
-
-
-#### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
+# There are 5 datasets used in analysis in total, all of which are provided by
+# the World Population Review for use. Each dataset can be manually downloaded
+# directly as a csv file. Data on incarceration rates by country can be downloaded
+# at https://worldpopulationreview.com/country-rankings/incarceration-rates-by-country,
+# data on poverty rates at https://worldpopulationreview.com/country-rankings/poverty-rate-by-country,
+# education rates at https://worldpopulationreview.com/country-rankings/education-rankings-by-country,
+# unemployment rates at https://worldpopulationreview.com/country-rankings/unemployment-by-country,
+# and crime rates at https://worldpopulationreview.com/country-rankings/crime-rate-by-country.
 
          

### 2024-04-16 scripts/02-data_cleaning.R
Cleaned data and saved file.
Merged all 5 raw datasets into one and cleaned it for use in the model and analysis. Then saved the file into analysis_data

@@ -1,44 +1,62 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Merges the 5 datasets listing each of the variables into one dataset 
+# and cleans them.
+# Author: Abbass Sleiman
+# Date: 15 April 2024
+# Contact: abbass.sleiman@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: Raw datasets must be downloaded as per the instructions in the 
+# 01-download_data.R script.
 
 #### Workspace setup ####
 library(tidyverse)
+library(janitor)
+library(here)
+library(dplyr)
+
+#### Read data ####
+raw_incarceration_data <- 
+  read_csv("data/raw_data/incarceration-rates-by-country-2024.csv")
+
+raw_crime_data <-
+  read_csv("data/raw_data/crime-rate-by-country-2024.csv")
+
+raw_education_data <-
+  read_csv("data/raw_data/education-rankings-by-country-2024.csv")
+
+raw_poverty_data <-
+  read_csv("data/raw_data/poverty-rate-by-country-2024.csv")
+
+raw_unemployment_data <-
+  read_csv("data/raw_data/unemployment-by-country-2024.csv")
+
+#### Merge data ####
+raw_merged_data <- inner_join(raw_incarceration_data, raw_crime_data, by = "country") |>
+  inner_join(raw_education_data, by = "country") |>
+  inner_join(raw_poverty_data, by = "country") |>
+  inner_join(raw_unemployment_data, by = "country")
 
 #### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
-
 cleaned_data <-
-  raw_data |>
+  raw_merged_data |>
   janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
+  rename(
[…]

### 2024-04-16 scripts/00-simulate_data.R
Simulated data and tested data.
Created a simulated dataset and conducted 5 tests to ensure that it works as expected. Moreover, wrote 4 tests to ensure proper functionality of the real cleaned dataset.

@@ -1,19 +1,84 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates 196 nations with randomly assigned incarceration rates,
+# poverty rates, crime rates, unemployment rates, and educational rankings. This
+# script also runs 5 tests to ensure proper functionality of the simulated dataset.
+# Author: Abbass Sleiman
+# Date: 16 April 2024
+# Contact: abbass.sleiman@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: None.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(9)
+
+# Number of countries
+num_countries <- 196
+
+# Generate educational ranking (unique integer values from 1 to 196)
+educational_ranking <- sample(1:196, num_countries)
+
+# Generate random values for other variables
+
+# Mean incarceration rate of 200 per 100k, with standard deviation of 50, rounded to 1 decimal place
+incarceration_rate <- round(abs(rnorm(num_countries, mean = 200, sd = 50)), 1)  
+
+# Mean poverty rate of 15%, with standard deviation of 5, rounded to 1 decimal place
+poverty_rate <- round(abs(rnorm(num_countries, mean = 15, sd = 5)), 1)    
+
+# Mean crime rate of 40 per 100k, with standard deviation of 20, rounded to 1 decimal place
+crime_rate <- round(abs(rnorm(num_countries, mean = 40, sd = 20)), 1) 
+
+# Mean unemployment rate of 8%, with standard deviation of 4, rounded to 1 decimal place
+unemployment_rate <- round(abs(rnorm(num_countries, mean = 8, sd = 4)), 1)      
+
+# Create dataset
+simulated_data <- data.frame(
+  Country = paste0("Country ", 1:num_countries),
+  Incarceration_Rate = incarceration_rate,
+  Poverty_Rate = poverty_rate,
+  Crime_Rate = crime_rate,
+  Unemployment_Rate = unemployment_rate,
[…]

### 2024-04-16 scripts/03-test_data.R

@@ -1,15 +1,46 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Conducts 4 tests to verify that the cleaned dataset works as expected.
+# Each of the 5 tests is outlined below.
+# Author: Abbass Sleiman
+# Date: 16 April 2024
+# Contact: abbass.sleiman@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: This script requires the user to have downloaded the data as per
+# the instructions outlined in "01-download_data.R", as well as run the script
+# "02-data_cleaning.R" in order to have access to the finalized cleaned data.
 
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(here)
 
-#### Test data ####
+#### Read data ####
+cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
+
+#### Test Data ####
+# Test 1: Check if all values are non-negative
+test_non_negative <- all(cleaned_data$incarceration_rate >= 0) &&
+  all(cleaned_data$poverty_rate >= 0) &&
+  all(cleaned_data$crime_rate >= 0) &&
+  all(cleaned_data$unemployment_rate >= 0)
+
+# Test 2: Check if there are 105 unique countries
+test_num_countries <- length(unique(cleaned_data$country)) == 105
+
+# Test 3: Check if there are no repeats in the educational ranking
+test_unique_educational_ranking <- length(unique(cleaned_data$education_rank)) == 105
+
+# Test 4: Check if each column is of the correct class
+test_column_classes <- class(cleaned_data$country) == "character" &&
+  class(cleaned_data$incarceration_rate) == "numeric" &&
+  class(cleaned_data$poverty_rate) == "numeric" &&
+  class(cleaned_data$crime_rate) == "numeric" &&
+  class(cleaned_data$unemployment_rate) == "numeric" &&
+  class(cleaned_data$education_rank) == "numeric"
+
+# Print test results
+print(paste("Test 1 (All values are non-negative):", test_non_negative))
+print(paste("Test 2 (There are 105 unique countries):", test_num_countries))
[…]

### 2024-04-16 scripts/02-data_cleaning.R
Updated dataset.
Changed the education ranking column to instead categorize educational rankings by deciles based on the number of countries that exist in the dataset (105).

@@ -46,17 +46,26 @@ cleaned_data <-
     crime_rate = crime_rate_by_country_crime_index,
     education_rank = education_rankings_by_country_wt20rank2024,
     unemployment_rate = unemployment_rate_world_bank
-  ) |>
+  )
+
+# Define the decile boundaries
+decile_boundaries <- c(1, 16, 28, 41, 58, 78, 101, 119, 139, 175, 198)
+
+# Create a new column in the dataset to store the education rank categories
+cleaned_data <- cleaned_data |>
+  mutate(education_category = cut(education_rank, breaks = decile_boundaries, 
+                                  labels = c("1-16", "17-28", "29-41", "42-58", 
+                                             "59-78", "79-101", "102-119", 
+                                             "120-139", "140-175", "176-198"))) |>
   select(
     country,
     incarceration_rate,
     poverty_rate,
     crime_rate,
     unemployment_rate,
-    education_rank
+    education_category
   ) |>
   na.omit()
-  
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/incarceration_data.csv")

### 2024-04-16 scripts/03-test_data.R

@@ -27,8 +27,8 @@ test_non_negative <- all(cleaned_data$incarceration_rate >= 0) &&
 # Test 2: Check if there are 105 unique countries
 test_num_countries <- length(unique(cleaned_data$country)) == 105
 
-# Test 3: Check if there are no repeats in the educational ranking
-test_unique_educational_ranking <- length(unique(cleaned_data$education_rank)) == 105
+# Test 3: Check if there are exactly 10 unique categories in the education category
+test_unique_categories <- length(unique(cleaned_data$education_category)) == 10
 
 # Test 4: Check if each column is of the correct class
 test_column_classes <- class(cleaned_data$country) == "character" &&
@@ -36,11 +36,11 @@ test_column_classes <- class(cleaned_data$country) == "character" &&
   class(cleaned_data$poverty_rate) == "numeric" &&
   class(cleaned_data$crime_rate) == "numeric" &&
   class(cleaned_data$unemployment_rate) == "numeric" &&
-  class(cleaned_data$education_rank) == "numeric"
+  class(cleaned_data$education_category) == "character"
 
 # Print test results
 print(paste("Test 1 (All values are non-negative):", test_non_negative))
 print(paste("Test 2 (There are 105 unique countries):", test_num_countries))
-print(paste("Test 3 (There are no repeats in the educational ranking):", test_unique_educational_ranking))
+print(paste("Test 3 (There are exactly 10 unique categories in the education category):", test_unique_categories))
 print(paste("Test 4 (Each column is of the correct class):", test_column_classes))
 

### 2024-04-16 scripts/04-model.R

@@ -1,37 +1,41 @@
 #### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Uses a Bayesian linear regression model to model and predict the effect
+# of poverty rate, crime rate, educational ranking, and unemployment on incarceration
+# rates.
+# Author: Abbass Sleiman
+# Date: 16 April 2024
+# Contact: abbass.sleiman@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
-
+# Pre-requisites: This script requires the user to have access to the cleaned 
+# dataset which can be accessed by having downloaded the raw data as per the 
+# instructions outlined in "01-download_data.R" and then having run the script
+# "02-data_cleaning.R".
 
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
 
 #### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
+cleaned_data <- read_csv("data/analysis_data/incarceration_data.csv")
 
 ### Model data ####
-first_model <-
+incarceration_model <-
   stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
+    formula = incarceration_rate ~ poverty_rate + crime_rate + educational_ranking + 
+      unemployment_rate + crime_rate * poverty_rate,
+    data = cleaned_data,
     family = gaussian(),
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
+    seed = 9
   )
 
 
 #### Save model ####
 saveRDS(
-  first_model,
-  file = "models/first_model.rds"
+  incarceration_model,
+  file = here("models/incarceration_model.rds")
 )
 
 

### 2024-04-17 scripts/02-data_cleaning.R
Updated data once more and ran a test model

@@ -63,9 +63,14 @@ cleaned_data <- cleaned_data |>
     poverty_rate,
     crime_rate,
     unemployment_rate,
+    education_rank,
     education_category
   ) |>
   na.omit()
+
+# Convert education_category into a factor variable
+cleaned_data <- cleaned_data |>
+  mutate(education_category = factor(education_category))
 
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/incarceration_data.csv")

### 2024-04-17 scripts/03-test_data.R

@@ -17,6 +17,10 @@ library(here)
 #### Read data ####
 cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
 
+#### Convert education variable into factor ####
+cleaned_data <- cleaned_data |>
+  mutate(education_category = factor(education_category))
+
 #### Test Data ####
 # Test 1: Check if all values are non-negative
 test_non_negative <- all(cleaned_data$incarceration_rate >= 0) &&
@@ -36,7 +40,7 @@ test_column_classes <- class(cleaned_data$country) == "character" &&
   class(cleaned_data$poverty_rate) == "numeric" &&
   class(cleaned_data$crime_rate) == "numeric" &&
   class(cleaned_data$unemployment_rate) == "numeric" &&
-  class(cleaned_data$education_category) == "character"
+  class(cleaned_data$education_category) == "factor"
 
 # Print test results
 print(paste("Test 1 (All values are non-negative):", test_non_negative))

### 2024-04-17 scripts/04-model.R

@@ -14,23 +14,58 @@
 #### Workspace setup ####
 library(tidyverse)
 library(rstanarm)
+library(modelsummary)
 
 #### Read data ####
 cleaned_data <- read_csv("data/analysis_data/incarceration_data.csv")
 
+#### Convert education to a factor variable ####
+cleaned_data <- cleaned_data |>
+  mutate(education_category = factor(education_category))
+
+#### Make the 176-198 the reference category ####
+cleaned_data$education_category <- relevel(cleaned_data$education_category, ref = "176-198")
+
 ### Model data ####
 incarceration_model <-
   stan_glm(
-    formula = incarceration_rate ~ poverty_rate + crime_rate + educational_ranking + 
-      unemployment_rate + crime_rate * poverty_rate,
+    formula = incarceration_rate ~ poverty_rate + crime_rate + education_category  + 
+      unemployment_rate + education_category * poverty_rate,
     data = cleaned_data,
     family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior = c(
+      # Prior specification for poverty_rate coefficient
+      normal(location = 0, scale = 2.5, autoscale = TRUE),
+      
+      # Prior specification for crime_rate coefficient
+      normal(location = 0, scale = 2.5, autoscale = TRUE),
+      
+      # Prior specification for education category coefficient
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 1-16
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 17-28
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 29-41
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 42-58
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 59-78
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 79-101
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 102-119
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 120-139
+      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 140-175
+      
+      # Prior specification for unemployment_rate coefficient
[…]

### 2024-04-17 paper/paper.qmd
Began data section
Began data section by adding some background information

@@
 
+This paper utilizes 5 datasets, all provided by the World Population Review and each providing information on one of the 5 factors involved in this study.
+@sec-incarceration-data features the most recently available information on incarcerations and incarceration rates of 219 countries and territories as of 2024.
+@sec-poverty-data features the data on the poverty rates of 164 nations.
+@sec-crime-data
+
+## Incarceration Rate Dataset {#sec-incarceration-data}
+
+## Poverty Rate Dataset {#sec-poverty-data}
+
+## Crime Rate Dataset {#sec-crime-data}
+
+## Unemployment Rate Dataset {#sec-unemployment-data}
+
+## Education Ranking Dataset {#sec-education-data}
+
 # Model {#sec-model}

### 2024-04-17 scripts/01-download_data.R
Updated data cleaning.
Changed one of the datasets from crime index to a more objective measure by measuring total violent crime

@@ -14,6 +14,6 @@
 # data on poverty rates at https://worldpopulationreview.com/country-rankings/poverty-rate-by-country,
 # education rates at https://worldpopulationreview.com/country-rankings/education-rankings-by-country,
 # unemployment rates at https://worldpopulationreview.com/country-rankings/unemployment-by-country,
-# and crime rates at https://worldpopulationreview.com/country-rankings/crime-rate-by-country.
+# and violent crime rates at https://worldpopulationreview.com/country-rankings/violent-crime-rates-by-country.
 
          

### 2024-04-17 scripts/02-data_cleaning.R

@@ -18,8 +18,8 @@ library(dplyr)
 raw_incarceration_data <- 
   read_csv("data/raw_data/incarceration-rates-by-country-2024.csv")
 
-raw_crime_data <-
-  read_csv("data/raw_data/crime-rate-by-country-2024.csv")
+raw_violent_crime_data <-
+  read_csv("data/raw_data/violent-crime-rates-by-country-2024.csv")
 
 raw_education_data <-
   read_csv("data/raw_data/education-rankings-by-country-2024.csv")
@@ -30,23 +30,28 @@ raw_poverty_data <-
 raw_unemployment_data <-
   read_csv("data/raw_data/unemployment-by-country-2024.csv")
 
-#### Merge data ####
-raw_merged_data <- inner_join(raw_incarceration_data, raw_crime_data, by = "country") |>
+# Merge data
+raw_merged_data <- inner_join(raw_incarceration_data, raw_violent_crime_data, by = "country") |>
   inner_join(raw_education_data, by = "country") |>
   inner_join(raw_poverty_data, by = "country") |>
   inner_join(raw_unemployment_data, by = "country")
 
-#### Clean data ####
+# Clean data
 cleaned_data <-
   raw_merged_data |>
   janitor::clean_names() |>
   rename(
     incarceration_rate = incarceration_rates_by_country_rate_per100k,
     poverty_rate = poverty_rate_value,
-    crime_rate = crime_rate_by_country_crime_index,
     education_rank = education_rankings_by_country_wt20rank2024,
     unemployment_rate = unemployment_rate_world_bank
   )
+
+# Calculate violent crime rate per country
+cleaned_data$violent_crime_rate <- rowSums(cleaned_data[, c("violent_crime_rates_by_country_hom_rate", 
+                                                                  "violent_crime_rates_by_country_assault", 
+                                                                  "violent_crime_rates_by_country_sex_violence", 
+                                                                  "violent_crime_rates_by_country_robbery")], na.rm = TRUE)
 
 # Define the decile boundaries
 decile_boundaries <- c(1, 16, 28, 41, 58, 78, 101, 119, 139, 175, 198)
@@ -61,11 +66,11 @@ cleaned_data <- cleaned_data |>
     country,
     incarceration_rate,
     poverty_rate,
-    crime_rate,
     unemployment_rate,
     education_rank,
-    education_category
-  ) |>
+    education_category,
+    violent_crime_rate
+    ) |>
   na.omit()
 
 # Convert education_category into a factor variable

### 2024-04-17 scripts/02-data_cleaning.R
Added one more variable to the total violent crime.
Added kidnapping as one more metric to measure total violent crimes in a nation

@@ -50,7 +50,8 @@ cleaned_data <-
 # Calculate violent crime rate per country
 cleaned_data$violent_crime_rate <- rowSums(cleaned_data[, c("violent_crime_rates_by_country_hom_rate", 
                                                                   "violent_crime_rates_by_country_assault", 
-                                                                  "violent_crime_rates_by_country_sex_violence", 
+                                                                  "violent_crime_rates_by_country_sex_violence",
+                                                                  "violent_crime_rates_by_country_kidnap",
                                                                   "violent_crime_rates_by_country_robbery")], na.rm = TRUE)
 
 # Define the decile boundaries

### 2024-04-17 scripts/02-data_cleaning.R
Updated deciles.
Updated the deciles to reflect the new number of observations, thereby requiring a change in the defining of categories.

@@ -55,14 +55,14 @@ cleaned_data$violent_crime_rate <- rowSums(cleaned_data[, c("violent_crime_rates
                                                                   "violent_crime_rates_by_country_robbery")], na.rm = TRUE)
 
 # Define the decile boundaries
-decile_boundaries <- c(1, 16, 28, 41, 58, 78, 101, 119, 139, 175, 198)
+decile_boundaries <- c(1, 19, 34, 52, 71, 92, 110, 125, 147, 175, 198)
 
 # Create a new column in the dataset to store the education rank categories
 cleaned_data <- cleaned_data |>
   mutate(education_category = cut(education_rank, breaks = decile_boundaries, 
-                                  labels = c("1-16", "17-28", "29-41", "42-58", 
-                                             "59-78", "79-101", "102-119", 
-                                             "120-139", "140-175", "176-198"))) |>
+                                  labels = c("1-19", "20-34", "35-52", "53-71", 
+                                             "72-92", "93-110", "111-125", 
+                                             "126-147", "148-175", "176-198"))) |>
   select(
     country,
     incarceration_rate,

### 2024-04-17 scripts/03-test_data.R
Updated tests.
Updated tests to reflect the new cleaned dataset.

@@ -1,6 +1,6 @@
 #### Preamble ####
 # Purpose: Conducts 4 tests to verify that the cleaned dataset works as expected.
-# Each of the 5 tests is outlined below.
+# Each of the 4 tests is outlined below.
 # Author: Abbass Sleiman
 # Date: 16 April 2024
 # Contact: abbass.sleiman@mail.utoronto.ca
@@ -25,11 +25,11 @@ cleaned_data <- cleaned_data |>
 # Test 1: Check if all values are non-negative
 test_non_negative <- all(cleaned_data$incarceration_rate >= 0) &&
   all(cleaned_data$poverty_rate >= 0) &&
-  all(cleaned_data$crime_rate >= 0) &&
+  all(cleaned_data$violent_crime_rate >= 0) &&
   all(cleaned_data$unemployment_rate >= 0)
 
-# Test 2: Check if there are 105 unique countries
-test_num_countries <- length(unique(cleaned_data$country)) == 105
+# Test 2: Check if there are 134 unique countries
+test_num_countries <- length(unique(cleaned_data$country)) == 134
 
 # Test 3: Check if there are exactly 10 unique categories in the education category
 test_unique_categories <- length(unique(cleaned_data$education_category)) == 10
@@ -38,13 +38,13 @@ test_unique_categories <- length(unique(cleaned_data$education_category)) == 10
 test_column_classes <- class(cleaned_data$country) == "character" &&
   class(cleaned_data$incarceration_rate) == "numeric" &&
   class(cleaned_data$poverty_rate) == "numeric" &&
-  class(cleaned_data$crime_rate) == "numeric" &&
+  class(cleaned_data$violent_crime_rate) == "numeric" &&
   class(cleaned_data$unemployment_rate) == "numeric" &&
   class(cleaned_data$education_category) == "factor"
 
 # Print test results
 print(paste("Test 1 (All values are non-negative):", test_non_negative))
-print(paste("Test 2 (There are 105 unique countries):", test_num_countries))
+print(paste("Test 2 (There are 134 unique countries):", test_num_countries))
 print(paste("Test 3 (There are exactly 10 unique categories in the education category):", test_unique_categories))
 print(paste("Test 4 (Each column is of the correct class):", test_column_classes))
 

### 2024-04-17 scripts/00-simulate_data.R
Updated simulated data

@@ -30,8 +30,8 @@ incarceration_rate <- round(abs(rnorm(num_countries, mean = 200, sd = 50)), 1)
 # Mean poverty rate of 15%, with standard deviation of 5, rounded to 1 decimal place
 poverty_rate <- round(abs(rnorm(num_countries, mean = 15, sd = 5)), 1)    
 
-# Mean crime rate of 40 per 100k, with standard deviation of 20, rounded to 1 decimal place
-crime_rate <- round(abs(rnorm(num_countries, mean = 40, sd = 20)), 1) 
+# Mean violent crime rate of 200 per 100k, with standard deviation of 200, rounded to 1 decimal place
+violent_crime_rate <- round(abs(rnorm(num_countries, mean = 200, sd = 200)), 1) 
 
 # Mean unemployment rate of 8%, with standard deviation of 4, rounded to 1 decimal place
 unemployment_rate <- round(abs(rnorm(num_countries, mean = 8, sd = 4)), 1)      
@@ -41,7 +41,7 @@ simulated_data <- data.frame(
   Country = paste0("Country ", 1:num_countries),
   Incarceration_Rate = incarceration_rate,
   Poverty_Rate = poverty_rate,
-  Crime_Rate = crime_rate,
+  Violent_Crime_Rate = violent_crime_rate,
   Unemployment_Rate = unemployment_rate,
   Educational_Ranking = educational_ranking
 )
@@ -50,7 +50,7 @@ simulated_data <- data.frame(
 # Test 1: Check if all values are non-negative
 test_non_negative <- all(simulated_data$Incarceration_Rate >= 0) &&
   all(simulated_data$Poverty_Rate >= 0) &&
-  all(simulated_data$Crime_Rate >= 0) &&
+  all(simulated_data$Violent_Crime_Rate >= 0) &&
   all(simulated_data$Unemployment_Rate >= 0)
 
 # Test 2: Check if there are 196 unique countries
@@ -63,14 +63,14 @@ test_unique_educational_ranking <- length(unique(simulated_data$Educational_Rank
 test_column_classes <- class(simulated_data$Country) == "character" &&
   class(simulated_data$Incarceration_Rate) == "numeric" &&
   class(simulated_data$Poverty_Rate) == "numeric" &&
-  class(simulated_data$Crime_Rate) == "numeric" &&
+  class(simulated_data$Violent_Crime_Rate) == "numeric" &&
[…]

### 2024-04-17 paper/paper.qmd
test
Updated data section.
Added information on all 5 datasets including snippets of each of the datasets. Also made a first trial model

@@
 
+# Package loading
 library(tidyverse)
+library(here)
+library(ggplot2)
+library(knitr)
+
+# Read data
+raw_incarceration_data <- read_csv(here("data/raw_data/incarceration-rates-by-country-2024.csv"))
+raw_poverty_data <- read_csv(here("data/raw_data/poverty-rate-by-country-2024.csv"))
+raw_violent_crime_data <- read_csv(here("data/raw_data/violent-crime-rates-by-country-2024.csv"))
+raw_unemployment_data <- read_csv(here("data/raw_data/unemployment-by-country-2024.csv"))
+raw_education_data <- read_csv(here("data/raw_data/education-rankings-by-country-2024.csv"))
+
+cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
 ```
@@
 
-This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Crime Rate by Country 2024".
+This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Violent Crime Rates by Country 2024".
 By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that [...]
@@
 This paper utilizes 5 datasets, all provided by the World Population Review and each providing information on one of the 5 factors involved in this study.
-@sec-incarceration-data features the most recently available information on incarcerations and incarceration rates of 219 countries and territories as of 2024.
-@sec-poverty-data features the data on the poverty rates of 164 nations.
-@sec-crime-data
+The incarceration rate dataset [@citeIncarcerationRate], @sec-incarceration-data, features the most recently available information on incarcerations and incarceration rates of 219 countries and territories as of 2024.
+The poverty rate dataset [@citePovertyRate], @sec-poverty-data, features the data on the poverty rates of 164 nations.
+The violent crime rate dataset [@citeViolentCrimeRate], @sec-violent-crime-data, features data on the number of various types of violent crimes in each of 200 nations and territories per 100k members of the population.
+The unemployment rate dataset [@citeUnemploymentRate], @sec-unemployment-data, features data on the unemployment rate in each of 226 countries and territories.
+Finally, the education ranking dataset [@citeEducationRanking], @sec-education-data, contains data on the literacy rate and education rank of each of 207 countries and territories, accurate as of 2024.
+
+All data analysis was done through R [@citeR] with the aid of the following packages:
 
@@
 
+The raw incarceration rate dataset contains information on the incarceration rate (per 100k population), total incarcerated, percentage of incarcerated that are female, and percentage of incarcerated that are male for 219 countries and territories.
+A snippet of the top 10 countries ordered by incarceration rate can be seen in @tbl-incarceration-data, showcasing their rates of incarceration and total number of individuals incarcerated.
+It is important to note that, whilst the data provided is the most recent available for each nation, the data for each nation need not be accurate as of 2024 as varying nations may have data from different years.
+
+```{r}
+#| label: tbl-incarceration-data
+#| tbl-cap: Top 10 countries by incarceration rate
+#| echo: false
+
+raw_incarceration_data |>
+  arrange(desc(incarcerationRatesByCountry_ratePer100k)) |>
+  head(10) |>
+  select(country, incarcerationRatesByCountry_ratePer100k, incarcerationRatesByCountry_totIncarcerated) |>
+  kable(
+    col.names = c("Country", "Incarceration Rate per 100k Population", "Total Incarcerated")
+  )
+```
+
 ## Poverty Rate Dataset {#sec-poverty-data}
 
-## Crime Rate Dataset {#sec-crime-data}
+The poverty rate dataset contains information on the poverty rates in 164 nations.
+In particular, it provides information on the percentage of the population living on less than 10 USD per day, less than 3.65 USD per day, the poverty rate, and the year the data was collected.
+Here, poverty rate is measured as the percentage of a given population living below that particular nation's respective poverty line which is estimated based on population-weighted subgroup estimates from household surveys [@citePovertyRate].
+This is an important clarification to make as it ensures that a nation's poverty rate is relative to the living costs in that particular nation, in turn reducing biases that could arise as a result of the vastly differing costs of living across various nations.
+This ensures that, in the analysis that shall follow, we are able to more accurately take into account income inequality relative to each particular nation.
+The top 5 countries ordered by poverty rate are showcased in @tbl-poverty-data, showcasing information on the poverty rate and year of data collection for each of the 5 countries.
+
+```{r}
+#| label: tbl-poverty-data
+#| tbl-cap: Top 5 countries by poverty rate
+#| echo: false
+
+raw_poverty_data |>
+  arrange(desc(PovertyRateValue)) |>
+  head(5) |>
+  select(country, PovertyRateValue, PovertyRateValueDataYear) |>
+  kable(
+    col.names = c("Country", "Poverty Rate", "Year Data was Collected")
+  )
+```
+
+## Violent Crime Rate Dataset {#sec-violent-crime-data}
+
+The violent crime rate dataset contains violent crime statistics for 200 nations and territories across the globe.
+In particular, it contains data on the homicide rate, serious assault rate, sexual violence rate, kidnapping rate, and robbery rate for each of the nations, each of which is measured to be the number of incidents per 100k population.
+It also features information on the year each of the data points was collected.
+As with the other datasets, the year of data collection for each nation varies by nation, and thus some information is more recent than others.
+
+An important note to make regarding this information is that the definitions for each of these crimes can vary significantly from nation to nation.
+What one nation may deem to be a serious offence can be nothing more than a misdemeanor in another.
+As a result, though the data provided is based on each nation's own definition of a serious offence, the fact that we are still provided with a fairly well-rounded subset of violent crimes ranging from theft to murder will enable us to better understand what the overall rate of violent crimes per nation is comprised of.
+One should note that some countries have missing values for certain crime rates, indicated by an "NA".
+@tbl-violent-crime-data showcases a sample of 5 countries including information regarding the aforementioned crime rates for each.
+
+```{r}
+#| label: tbl-violent-crime-data
+#| tbl-cap: Sample of 5 countries and their respective rates of various violent crimes (per 100k population)
+#| echo: false
+
+raw_violent_crime_data |>
+  head(5) |>
+  select(country, violentCrimeRatesByCountry_homRate, violentCrimeRatesByCountry_assault, violentCrimeRatesByCountry_sexViolence, violentCrimeRatesByCountry_kidnap, violentCrimeRatesByCountry_robbery) |>
+  kable(
+    col.names = c("Country", "Homicide Rate", "Serious Assault Rate", "Sexual Violence Rate", "Kidnapping Rate", "Robbery Rate")
+  )
+```
 
@@
 
+The unemployment rate dataset provides data on the unemployment rates of 226 countries and territories across the globe according to 3 various sources: the World Bank, the International Labour Organization, and the CIA, as well as the respective years that each of these organizations collected the data.
+Here, unemployment is defined as the percentage of unemployed workers in the labour force, which includes those who do not currently work but are able to do so [@citeUnemploymentRate].
+It is important to note that, as with the other datasets, information on the nations' unemployment rates is not necessarily accurate as of 2024 and the date of collection varies from country to country.
+As the World Bank has information on every one of the countries and territories, this paper will utilize only data from that one source regarding unemployment rates.
+@tbl-unemployment-data showcases the top 5 countries by highest unemployment rates, according to the World Bank, as well as the year the data was collected.
+
+```{r}
+#| label: tbl-unemployment-data
+#| tbl-cap: Top 5 countries by highest unemployment rate (as per the World Bank)
+#| echo: false
+
+raw_unemployment_data |>
+  arrange(desc(UnemploymentRateWorldBank)) |>
+  head(5) |>
+  select(country, UnemploymentRateWorldBank,UnemploymentRateWorldBankDataYear) |>
+  kable(
+    col.names = c("Country", "Unemployment Rate", "Year Data was Collected")
+  )
+```
+
 ## Education Ranking Dataset {#sec-education-data}
+
+The education ranking dataset provides information on the literacy rate and education rank of 207 countries and territories.
+The dataset contains ranks assigned by various organizations, however the most recent ranking data is by the nonprofit World Top 20, who determine a nation's rank according to a number of factors including early childhood enrollment rates, high school graduation rates, primary test scores, and various other factors [@citeEducationCriteria].
+The top 5 countries according to the World Top 20 in 2024 are showcased in @tbl-education-data.
+
+```{r}
+#| label: tbl-education-data
+#| tbl-cap: Top 5 countries by education (World Top 20 (2024))
+#| echo: false
+
+raw_education_data |>
+  arrange(educationRankingsByCountry_WT20rank2024) |>
+  head(5) |>
+  select(country, educationRankingsByCountry_WT20rank2024) |>
+  kable(
+    col.names = c("Country", "Education Rank")
+  )
+```
+
+## Cleaned Dataset {#sec-cleaned-data}
 

### 2024-04-17 scripts/04-model.R

@@ -29,7 +29,7 @@ cleaned_data$education_category <- relevel(cleaned_data$education_category, ref 
 ### Model data ####
 incarceration_model <-
   stan_glm(
-    formula = incarceration_rate ~ poverty_rate + crime_rate + education_category  + 
+    formula = incarceration_rate ~ poverty_rate + violent_crime_rate + education_category  + 
       unemployment_rate + education_category * poverty_rate,
     data = cleaned_data,
     family = gaussian(),

### 2024-04-17 paper/paper.qmd
Updated data, finished data section, and created model.
Removed an unneeded variable from the data, finalized data section in quarto, and updated model

@@
 
+The finalized cleaned dataset used in the analysis that will follow was created by merging the aforementioned 5 datasets into one.
+In particular, only countries that were present in all 5 datasets are present in the cleaned dataset, bringing the total number of nations included in this study to 134.
+Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its national poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category.
+Note that a nation's violent crime rate and educational ranking category were not explicit data that were provided as part of the 5 raw datasets.
+Instead, these were manually constructed and appended to the final dataset during the cleaning process.
+
+In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation.
+It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the sum.
+As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, it was instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking.
+This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective.
+@tbl-cleaned-data
+
+```{r}
+#| label: tbl-cleaned-data
+#| tbl-cap: Top 5 countries by incarceration rate in the finalized cleaned dataset
+#| echo: false
+
+cleaned_data |>
+  arrange(desc(incarceration_rate)) |>
+  head(5) |>
+  kable(
+        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate")
+  )
+```
+
+To gain a better understanding of the distribution of incarceration rates as a function of the independent variables, the figures below each showcase all observations, placing incarceration rates on the y-axis and independent variable on the x-axis.
+Moreover, we have chosen to highlight 5 particular nations in each of the figures, representing nations across the entire range of incarceration rates present in the dataset.
+In particular, we highlight El Salvador who has the highest incarceration rate of 1086, the United States with the 3rd highest incarceration rate of 531, Mexico with the 51st highest incarceration rate of 174, Iceland with the 5th lowest incarceration rate of 36, and Guinea-Bissau with the lowest incarceration rate of 31.
+Do note that these rankings are relative to the finalized dataset as opposed to the entire list of countries and territories found in the raw incarceration rate dataset.
+
+@fig-poverty-incarceration showcases all observations by poverty rate, highlighting the aforementioned countries as well.
+Similarly, @fig-unemployment-incarceration showcases all observations by unemployment rate and @fig-violent-crime-incarceration by violent crime rates.
+Lastly, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories.
+Note that Iceland is in the "1-19" category, the United States is in the "20-34" category, Mexico is in the "35-52" category, El Salvador is in the "111-125" category, and Guinea-Bissau is in the "176-198" category.
+
+```{r}
+#| label: fig-poverty-incarceration
+#| fig-cap: Scatter plot of incarceration rates by poverty rates
+#| echo: false
+#| fig-pos: H
+
+ggplot(cleaned_data, aes(x = poverty_rate, y = incarceration_rate)) +
+  geom_point(color = "darkgrey") +
+  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
+             aes(color = country), size = 1.5) +  # Highlight specific countries
+  theme_classic() +
+  labs(x = "Poverty Rate", y = "Incarceration Rate")
+```
+
+```{r}
+#| label: fig-unemployment-incarceration
+#| fig-cap: Scatter plot of incarceration rates by unemployment rates
+#| echo: false
+#| fig-pos: H
+
+ggplot(cleaned_data, aes(x = unemployment_rate, y = incarceration_rate)) +
+  geom_point(color = "darkgrey") +
+  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
+             aes(color = country), size = 1.5) +  # Highlight specific countries
+  theme_classic() +
+  labs(x = "Unemployment Rate", y = "Incarceration Rate")
+```
+
+```{r}
+#| label: fig-violent-crime-incarceration
+#| fig-cap: Scatter plot of incarceration rates by violent crime rates
+#| echo: false
+#| fig-pos: H
+
+ggplot(cleaned_data, aes(x = violent_crime_rate, y = incarceration_rate)) +
+  geom_point(color = "darkgrey") +
+  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
+             aes(color = country), size = 1.5) +  # Highlight specific countries
+  theme_classic() +
+  labs(x = "Violent Crime Rate", y = "Incarceration Rate")
+```
+
+```{r}
+#| label: fig-education-incarceration
+#| fig-cap: Bar chart showcasing mean incarceration rate for each educational ranking category
+#| echo: false
+#| fig-pos: H
+
+# Define the levels in the desired order
+education_levels <- c("1-19", "20-34", "35-52", "53-71", "72-92", 
+                      "93-110", "111-125", "126-147", "148-175", "176-198")
+
+# Convert education_category to a factor with specified levels
+cleaned_data$education_category <- factor(cleaned_data$education_category, levels = education_levels)
+
+# Recalculate the mean incarceration rate for each educational category
+mean_incarceration <- cleaned_data |>
+  group_by(education_category) |>
+  summarize(mean_incarceration_rate = mean(incarceration_rate, na.rm = TRUE))
+
+# Create the bar chart
+ggplot(mean_incarceration, aes(x = education_category, y = mean_incarceration_rate)) +
+  geom_bar(stat = "identity", fill = "skyblue") +
+  theme_classic() +
+  labs(x = "Educational Ranking Category", y = "Mean Incarceration Rate") +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for better readability
+
+```
+
+To gain a better understanding of the underlying data, @tbl-summary-stats provides information on the overall mean and standard deviation of each of the variables present in the cleaned dataset.
+
+```{r}
+#| label: tbl-summary-stats
+#| tbl-cap: Mean and standard deviation for each variable
+#| echo: false
+#| fig-pos: H
+
+# Calculate summary statistics
+mean_incarceration <- round(mean(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
+sd_incarceration <- round(sd(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
+mean_poverty <- round(mean(cleaned_data$poverty_rate, na.rm = TRUE), 2)
+sd_poverty <- round(sd(cleaned_data$poverty_rate, na.rm = TRUE), 2)
+mean_unemployment <- round(mean(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
+sd_unemployment <- round(sd(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
+mean_violent <- round(mean(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
+sd_violent <- round(sd(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
+
+# Create a data frame manually
+data <- data.frame(
+  Variable = c("Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Violent Crime Rate"),
+  Mean = c(mean_incarceration, mean_poverty, mean_unemployment, mean_violent),
+  SD = c(sd_incarceration, sd_poverty, sd_unemployment, sd_violent)
+)
+
+# Print the data frame as a table using kable
+kable(data, col.names = c("Variable", "Mean", "SD"))
+```
+
 # Model {#sec-model}
@@
 # Results {#sec-results}
+Make note that poorer nations could also be incapable of housing more individuals in jails (use kenya as case study. 284% jail capacity)
 

### 2024-04-17 scripts/02-data_cleaning.R

@@ -68,7 +68,6 @@ cleaned_data <- cleaned_data |>
     incarceration_rate,
     poverty_rate,
     unemployment_rate,
-    education_rank,
     education_category,
     violent_crime_rate
     ) |>

### 2024-04-17 scripts/04-model.R

@@ -30,7 +30,7 @@ cleaned_data$education_category <- relevel(cleaned_data$education_category, ref 
 incarceration_model <-
   stan_glm(
     formula = incarceration_rate ~ poverty_rate + violent_crime_rate + education_category  + 
-      unemployment_rate + education_category * poverty_rate,
+      unemployment_rate,
     data = cleaned_data,
     family = gaussian(),
     prior = c(

### 2024-04-18 paper/paper.qmd
Attempted to save data as parquet

@@
 
+As it stands, by adhering to the figures above, there does not appear to be a strong correlation between any one of the variables and incarceration rates.
+In particular, poverty rates and unemployment rates do not showcase any particular trend in any direction.
+Violent crime on the other hand, does appear to show a somewhat positive correlation with incarceration rates.
+However what is interesting to note is that the 5 highlighted nations spanning the entire range of incarceration rates do not have a significant range of violent crime rates, at least relative to the overall range of violent crime rates in the entire dataset.
+The educational ranking category appears to have a negative relationship with incarceration rates, where a higher rank, i.e.
+a better education, seems to correspond with a lower average incarceration rate.
+However, somewhat interesting to note is the drop in mean incarceration rates for categories past the peak in the "111-125" ranking category.
+As such, given the fairly minimal information provided by the graphical and tabular figures above regarding the relationship between the variables of interest and a nation's incarceration rate, we employ a model in @sec-model to come to a more conclusive result.
+
 # Model {#sec-model}
+
+As the purpose of this paper is the investigation of various socioeconomic factors and their impact on a nation's incarceration rate as a means of better understanding the United States' mass incarceration problem, we will employ a model that will better enable us to infer the relationship between these various factors and incarceration rates.
+Then using this information, I will attempt to predict what the United States' incarceration rate should be given its own individual data for each of the determinants of incarceration rate, and subsequently uncover whether they are a good estimator of incarceration rate, or whether the problem stems from a different facet.
 
@@
 
+The particular model that this paper will utilize is a Bayesian multiple linear regression model of incarceration rate as a function of the independent variables present in our analysis.
+In particular, the model is as follows:
+
+\begin{align} 
+y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
+\mu_i &= \beta_0 + \beta_1 poverty_{i} + \beta_2 crime_{i} + \beta_3 unemployment_{i} + \gamma_1 educ_{i} + ... + \gamma_9 educ_{i} \\
+\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_2 &\sim \mbox{Normal}(0, 2.5) \\
+\beta_3 &\sim \mbox{Normal}(0, 2.5) \\
+\gamma_i &\sim \mbox{Normal}(0, 2.5) \\
+\sigma &\sim \mbox{Exponential}(1)
+\end{align}
+
+In the above model:
+
+- $\mu_i$ is the predicted incarceration rate for a nation given a nation's poverty rate, violent crime rate, unemployment rate, and educational ranking category.
+- $\beta_0$ is the coefficient for the intercept.
+- $\beta_1$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit, i.e. one percentage point, increase in the nation's poverty rate.
+- $\beta_2$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit increase in the nation's violent crime rate.
+- $\beta_3$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit, i.e. one percentage point, increase in the nation's unemployment rate.
+- $\gamma_i$, for $i = 1, ..., 9$, represents the coefficient for the predicted change in the incarceration rate of a nation given that it is in one of the educational ranking categories, "1-19", ..., "148-175", respectively for each $i$.
+Notice, this implies that the base intercept assumes that a nation is in the "176-198" educational ranking category.
+
+Note, we employ [@citerstanarm] in order to run the regression. Moreover, we make use of the default priors offered by `rstanarm`, however we allow for auto scaling.
+
 ### Model justification
+
+We expect
 

### 2024-04-18 scripts/02-data_cleaning.R

@@ -13,6 +13,7 @@ library(tidyverse)
 library(janitor)
 library(here)
 library(dplyr)
+library(arrow)
 
 #### Read data ####
 raw_incarceration_data <- 
@@ -78,4 +79,4 @@ cleaned_data <- cleaned_data |>
   mutate(education_category = factor(education_category))
 
 #### Save data ####
-write_csv(cleaned_data, "data/analysis_data/incarceration_data.csv")
+write_parquet(x = cleaned_data, sink = "data/analysis_data/incarceration_data.parquet")

### 2024-04-18 scripts/04-model.R

@@ -29,15 +29,18 @@ cleaned_data$education_category <- relevel(cleaned_data$education_category, ref 
 ### Model data ####
 incarceration_model <-
   stan_glm(
-    formula = incarceration_rate ~ poverty_rate + violent_crime_rate + education_category  + 
-      unemployment_rate,
+    formula = incarceration_rate ~ poverty_rate + violent_crime_rate + unemployment_rate  + 
+      education_category,
     data = cleaned_data,
     family = gaussian(),
     prior = c(
       # Prior specification for poverty_rate coefficient
       normal(location = 0, scale = 2.5, autoscale = TRUE),
       
-      # Prior specification for crime_rate coefficient
+      # Prior specification for violent_crime_rate coefficient
+      normal(location = 0, scale = 2.5, autoscale = TRUE),
+      
+      # Prior specification for unemployment_rate coefficient
       normal(location = 0, scale = 2.5, autoscale = TRUE),
       
       # Prior specification for education category coefficient
@@ -49,10 +52,9 @@ incarceration_model <-
       normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 79-101
       normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 102-119
       normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 120-139
-      normal(location = 0, scale = 2.5, autoscale = TRUE),  # Prior for category 140-175
+      normal(location = 0, scale = 2.5, autoscale = TRUE)  # Prior for category 140-175
       
-      # Prior specification for unemployment_rate coefficient
-      normal(location = 0, scale = 2.5, autoscale = TRUE)
+      
     ),
     prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
     prior_aux = exponential(rate = 1, autoscale = TRUE),

### 2024-04-18 scripts/02-data_cleaning.R
Undid parquet.
Reverted back to csv due to error in creating parquet. ReadME will document circumstances

@@ -13,7 +13,6 @@ library(tidyverse)
 library(janitor)
 library(here)
 library(dplyr)
-library(arrow)
 
 #### Read data ####
 raw_incarceration_data <- 
@@ -79,4 +78,4 @@ cleaned_data <- cleaned_data |>
   mutate(education_category = factor(education_category))
 
 #### Save data ####
-write_parquet(x = cleaned_data, sink = "data/analysis_data/incarceration_data.parquet")
+write_csv(cleaned_data, "data/analysis_data/incarceration_data.csv")

### 2024-04-18 paper/paper.qmd
Completed model section.
Added model section and justifications. Added some more literature to justify the choice of independent variables

@@
 
-We expect
+The above factors were each chosen due to the literature present regarding each of their impacts on incarceration.
+In particular, there is data to suggest that more impoverished individuals are more susceptible to being arrested, at least in the United States [@citePovertyImpact].
+Moreover, there is research to suggest that unemployment rates seem to correlate with incarceration rates, at least over extended periods of time in the US [@citeUnemploymentImpact].
+There is also literature that suggests that education significantly reduces the likelihood of incarceration, with claims that incarceration rates monotonically decline with education [@citeEducationImpact].
+As for the impact of violent crime rates on incarceration rates, most research generally tends to focus on the reverse, focusing on incarceration rates' impact on violent crime.
+As such, this paper seeks to explore the converse in order to gain a more well-rounded understanding of how both factors influence one another on the whole.
+
+A priori, we expect a linear relationship between each of these variables and a nation's incarceration rate.
+More specifically, we would expect that as the the level of violent crime, poverty, and unemployment rises, so too does the incarceration rate, as individuals may find themselves more likely to commit crimes as a means of survival, in turn increasing the amount of individuals that become imprisoned.
+On the contrary, we expect that more educated nations will generally see lower incarceration rates as better education is likely to be correlated with better standards of living and lower crime rates.
 
 # Results {#sec-results}
-Make note that poorer nations could also be incapable of housing more individuals in jails (use kenya as case study. 284% jail capacity)
 
 # Discussion {#sec-discussion}
+Make note that poorer nations could also be incapable of housing more individuals in jails (use kenya as case study. 284% jail capacity)
 

### 2024-04-18 paper/paper.qmd
Finished model and results section.
Completed the model section and include information regarding the results that can be derived as a result.

@@
 library(knitr)
+library(modelsummary)
+library(rstanarm)
+library(kableExtra)
 
@@
 cleaned_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
+
+# Read model
+analysis_model <- readRDS(file = here("models/incarceration_model.rds"))
+
 ```
@@
 #| echo: false
+#| tbl-pos: H
 
@@
   kable(
-    col.names = c("Country", "Incarceration Rate per 100k Population", "Total Incarcerated")
+    col.names = c("Country", "Incarceration Rate per 100k Population", "Total Incarcerated"), 
+    booktabs = TRUE
   )
@@
   kable(
-    col.names = c("Country", "Poverty Rate", "Year Data was Collected")
+    col.names = c("Country", "Poverty Rate", "Year Data was Collected"), 
+    booktabs = TRUE
   )
@@
   kable(
-    col.names = c("Country", "Homicide Rate", "Serious Assault Rate", "Sexual Violence Rate", "Kidnapping Rate", "Robbery Rate")
-  )
+    col.names = c("Country", "Homicide Rate", "Serious Assault Rate", "Sexual Violence Rate", "Kidnapping Rate", "Robbery Rate"), booktabs = TRUE
+  ) |>
+  column_spec(1, width = "1.8cm") |>
+  column_spec(2, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(3, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(4, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(5, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(6, width = "2.2cm")     # Adjust the width of the second column
 ```
@@
   kable(
-    col.names = c("Country", "Unemployment Rate", "Year Data was Collected")
+    col.names = c("Country", "Unemployment Rate", "Year Data was Collected"), 
+    booktabs = TRUE
   )
@@
   kable(
-    col.names = c("Country", "Education Rank")
+    col.names = c("Country", "Education Rank"), booktabs = TRUE
   )
@@
 In this study, a nation's violent crime rate is defined to be the sum of all the crime rates (per 100k population) available through the violent crime dataset in order to have one final value that represents the extent of violent crime in a nation.
-It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the sum.
-As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, it was instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking.
+It is important to note, however, that various nations had missing values for some of the violent crime statistics, and as a result, these values were simply treated as 0 for the sake of the summation.
+As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, we instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking.
 This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective.
-@tbl-cleaned-data
+@tbl-cleaned-data showcases the top 5 countries by incarceration rate in the cleaned dataset.
 
@@
   kable(
-        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate")
-  )
+        col.names = c("Country", "Incarceration Rate", "Poverty Rate", "Unemployment Rate", "Educational Ranking Category", "Violent Crime Rate"), booktabs = TRUE
+  ) |>
+  column_spec(1, width = "2.2cm") |>
+  column_spec(2, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(3, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(4, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(5, width = "2.2cm") |>  # Adjust the width of the first column
+  column_spec(6, width = "2.2cm")
 ```
@@
 
-@fig-poverty-incarceration showcases all observations by poverty rate, highlighting the aforementioned countries as well.
-Similarly, @fig-unemployment-incarceration showcases all observations by unemployment rate and @fig-violent-crime-incarceration by violent crime rates.
-Lastly, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories.
+@fig-variable-incarceration showcases all observations by poverty rate, unemployment rate, and violent crime rates, highlighting the aforementioned countries as well.
+On the other hand, @fig-education-incarceration showcases the mean incarceration rate for each of the 10 educational ranking categories.
 Note that Iceland is in the "1-19" category, the United States is in the "20-34" category, Mexico is in the "35-52" category, El Salvador is in the "111-125" category, and Guinea-Bissau is in the "176-198" category.
@@
 ```{r}
-#| label: fig-poverty-incarceration
-#| fig-cap: Scatter plot of incarceration rates by poverty rates
+#| label: fig-variable-incarceration
+#| fig-cap: Scatter plot of incarceration rates against each of the independent variables of interest.
+#| fig-subcap: ["Poverty Rate", "Unemployment Rate", "Violent Crime Rate"]
 #| echo: false
+#| layout-ncol: 2
 #| fig-pos: H
@@
   labs(x = "Poverty Rate", y = "Incarceration Rate")
-```
-
-```{r}
-#| label: fig-unemployment-incarceration
-#| fig-cap: Scatter plot of incarceration rates by unemployment rates
-#| echo: false
-#| fig-pos: H
 
@@
   labs(x = "Unemployment Rate", y = "Incarceration Rate")
-```
-
-```{r}
-#| label: fig-violent-crime-incarceration
-#| fig-cap: Scatter plot of incarceration rates by violent crime rates
-#| echo: false
-#| fig-pos: H
 
@@
 #| echo: false
-#| fig-pos: H
+#| tbl-pos: H
 
@@
 # Print the data frame as a table using kable
-kable(data, col.names = c("Variable", "Mean", "SD"))
+kable(data, col.names = c("Variable", "Mean", "SD"), booktabs = TRUE)
 ```
@@
 
-Note, we employ [@citerstanarm] in order to run the regression. Moreover, we make use of the default priors offered by `rstanarm`, however we allow for auto scaling.
+Note, we employ `rstanarm` [@citerstanarm] in order to run the regression. Moreover, we make use of the default priors offered by `rstanarm`, however we allow for auto scaling.
 
@@
 In particular, there is data to suggest that more impoverished individuals are more susceptible to being arrested, at least in the United States [@citePovertyImpact].
-Moreover, there is research to suggest that unemployment rates seem to correlate with incarceration rates, at least over extended periods of time in the US [@citeUnemploymentImpact].
+Moreover, there is research to suggest that unemployment rates seem to correlate with incarceration rates, at least over extended periods of time in the United States [@citeUnemploymentImpact].
 There is also literature that suggests that education significantly reduces the likelihood of incarceration, with claims that incarceration rates monotonically decline with education [@citeEducationImpact].
@@
 # Results {#sec-results}
+
+## Model Coefficients
+
+After running the regression based on the above model, we receive the following coefficient values as showcased in @tbl-incarcerationsummary.
+
+```{r}
+#| label: tbl-incarcerationsummary
+#| tbl-cap: Model summary of the predicted impact of crime, education, unemployment, and poverty on incarceration rates.
+#| echo: false
+#| warning: false
+#| message: false
+#| tbl-pos: H
+
+modelsummary(analysis_model)
+
+```
+
+We can see that, somewhat unexpectedly, increases in the poverty rate negatively correlate with incarceration rates, as do increases in unemployment, assuming all else held constant.
+On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant.
+As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the lowest category of education, "176-198", would see an increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing more consistent decreases near the highest ranked nations.
+
+## Actual vs. Predicted Incarceration Rates
+
+In order to get a better grasp of the predictive power of our model, we compare every nation's predicted incarceration rate, given their respective qualities, to their actual incarceration rate as showcased in @fig-actualvspredicted below.
+
+```{r}
+#| label: fig-actualvspredicted
+#| tbl-cap: Comparison of actual and predicted incarceration rates.
+#| echo: false
+#| fig-pos: H
+
+# Extract coefficient estimates
+coefficients <- coef(analysis_model)
+
+# Extract predictor variables (poverty, crime, unemployment, and education)
+predictors <- cleaned_data |>
+  select(poverty_rate, violent_crime_rate, unemployment_rate, education_category)
+
+# Calculate the predicted incarceration rate for each nation
+predicted_incarceration <- (
+  coefficients["(Intercept)"] +                      
+  coefficients["poverty_rate"] * predictors$poverty_rate +
+  coefficients["violent_crime_rate"] * predictors$violent_crime_rate +
+  coefficients["unemployment_rate"] * predictors$unemployment_rate +
+  coefficients["education_category1-19"] * as.numeric(predictors$education_category == "1-19") +
+  coefficients["education_category20-34"] * as.numeric(predictors$education_category == "20-34") +
+  coefficients["education_category35-52"] * as.numeric(predictors$education_category == "35-52") +
+  coefficients["education_category53-71"] * as.numeric(predictors$education_category == "53-71") +
+  coefficients["education_category72-92"] * as.numeric(predictors$education_category == "72-92") +
+  coefficients["education_category93-110"] * as.numeric(predictors$education_category == "93-110") +
+  coefficients["education_category111-125"] * as.numeric(predictors$education_category == "111-125") +
+  coefficients["education_category126-147"] * as.numeric(predictors$education_category == "126-147") +
+  coefficients["education_category148-175"] * as.numeric(predictors$education_category == "148-175")
+)
+
+# Add predicted incarceration rate as a new column in cleaned_data
+cleaned_data <- cleaned_data |>
+  mutate(predicted_incarceration_rate = predicted_incarceration)
+
+# Graph the resultant scatter plot
+ggplot(cleaned_data, aes(x = predicted_incarceration_rate, y = incarceration_rate)) +
+  geom_point(color = "darkgrey") +  # Add points for actual incarceration rates
+  geom_abline(intercept = 0, slope = 1, color = "black", linetype = "dashed") +  # Add y=x line
+  geom_point(data = cleaned_data |> filter(country %in% c("United States", "El Salvador", "Guinea-Bissau", "Iceland", "Mexico")),
+             aes(color = country), size = 1.5) +
+  labs(x = "Predicted Incarceration Rate", y = "Actual Incarceration Rate") + 
+  theme_classic()
+
+```
+
+In the above figure, each point represents a given nation's actual incarceration rate (on the y-axis), against their predicted incarceration rate (on the x-axis) based on our employed model.
+We include a dotted $y=x$ line for reference to showcase all the theoretical points where the actual incarceration rate of a nation equals the predicted incarceration rate.
+Note that any points below the dotted $y=x$ line signify nations whose predicted incarceration rates exceeds their actual incarceration rates, whereas points above the line signify nations whose actual exceeds their predicted.
+
+Examining the results on the whole, it appears that the model is able to somewhat accurately predict a nation's incarceration rate, whenever that nation's actual incarceration rate is between 100 and 300.
+As a nation's actual incarceration rises above 300, the model loses its predictive power as it appears to significantly underestimate the nation's incarceration rate.
+On the other hand, with nations whose actual incarceration rates are below 100, it appears that model generally tends to overestimate the nation's incarceration rate.
+
+We highlight the same 5 nations as done in @sec-data as they span the entire range of incarceration rates.
+We can see that, as observed, Iceland and Guinea-Bissau, who have the 5th lowest and lowest incarceration rates of 31 and 36, respectively, feature predicted incarceration rates that exceed their actual.
+El Salvador and the United States, who have the highest and 3rd highest incarceration rates of 1086 and 531, respectively, featured predicted incarceration rates far below their actual.
+On the other hand, Mexico, with an actual incarceration rate of 174, had a predicted incarceration rate almost exactly equal to its actual rate.
+Thus, these 5 nations effectively showcase where the model's predictive power is strongest - near the middle - and where it is weakest - on either extreme.
+
+We discuss the implications of these results in @sec-discussion, and in particular, how this helps in understanding the United States' issue of mass incarceration.
 
@@
 
+```{r}
+#| label: fig-ppcheck-postprior
+#| echo: false
+#| fig-pos: H
+#| message: false
+
+pp_check(analysis_model) + theme_classic()
+
+posterior_vs_prior(analysis_model) + theme_classic() + coord_flip()
+```
+
+```{r}
+#| label: fig-rhat
+#| echo: false
+#| layout-ncol: 2
+#| fig-pos: H
+
+plot(analysis_model, "rhat")
+```
 \newpage

### 2024-04-18 paper/paper.qmd
Updated model and model section.
Changed the reference educational category to that of "1-19" since there seemed to be an error with the prior and posterior choices of categories.

@@
 - $\beta_3$ is the coefficient for the predicted change in the incarceration rate of a nation given a one unit, i.e. one percentage point, increase in the nation's unemployment rate.
-- $\gamma_i$, for $i = 1, ..., 9$, represents the coefficient for the predicted change in the incarceration rate of a nation given that it is in one of the educational ranking categories, "1-19", ..., "148-175", respectively for each $i$.
-Notice, this implies that the base intercept assumes that a nation is in the "176-198" educational ranking category.
+- $\gamma_i$, for $i = 1, ..., 9$, represents the coefficient for the predicted change in the incarceration rate of a nation given that it is in one of the educational ranking categories, "20-34", ..., "176-198", respectively for each $i$.
+Notice, this implies that the base intercept assumes that a nation is in the "1-19" educational ranking category.
 
@@
 On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant.
-As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the lowest category of education, "176-198", would see an increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing more consistent decreases near the highest ranked nations.
+As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the highest category of education, "1-19", would see a much larger increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing smaller increase near the lowest ranked nations.
 
@@
   coefficients["unemployment_rate"] * predictors$unemployment_rate +
-  coefficients["education_category1-19"] * as.numeric(predictors$education_category == "1-19") +
   coefficients["education_category20-34"] * as.numeric(predictors$education_category == "20-34") +
@@
   coefficients["education_category126-147"] * as.numeric(predictors$education_category == "126-147") +
-  coefficients["education_category148-175"] * as.numeric(predictors$education_category == "148-175")
+  coefficients["education_category148-175"] * as.numeric(predictors$education_category == "148-175") +
+  coefficients["education_category176-198"] * as.numeric(predictors$education_category == "176-198") 
 )
@@
 #| echo: false
-#| layout-ncol: 2
 #| fig-pos: H

### 2024-04-18 scripts/04-model.R

@@ -24,7 +24,7 @@ cleaned_data <- cleaned_data |>
   mutate(education_category = factor(education_category))
 
 #### Make the 176-198 the reference category ####
-cleaned_data$education_category <- relevel(cleaned_data$education_category, ref = "176-198")
+cleaned_data$education_category <- relevel(cleaned_data$education_category, ref = "1-19")
 
 ### Model data ####
 incarceration_model <-

### 2024-04-18 paper/paper.qmd
Added 2 discussion points with references

@@
 
-To gain a better understanding of the underlying data, @tbl-summary-stats provides information on the overall mean and standard deviation of each of the variables present in the cleaned dataset.
+To gain a better understanding of the underlying data, @tbl-summary-stats provides information on the overall mean, median, and standard deviation of each of the variables present in the cleaned dataset.
 
@@
 #| label: tbl-summary-stats
-#| tbl-cap: Mean and standard deviation for each variable
+#| tbl-cap: Mean, median, and standard deviation for each variable
 #| echo: false
@@
 sd_incarceration <- round(sd(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
+median_incarceration <- round(median(cleaned_data$incarceration_rate, na.rm = TRUE), 2)
+
 mean_poverty <- round(mean(cleaned_data$poverty_rate, na.rm = TRUE), 2)
 sd_poverty <- round(sd(cleaned_data$poverty_rate, na.rm = TRUE), 2)
+median_poverty <- round(median(cleaned_data$poverty_rate, na.rm = TRUE), 2)
+
 mean_unemployment <- round(mean(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
 sd_unemployment <- round(sd(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
+median_unemployment <- round(median(cleaned_data$unemployment_rate, na.rm = TRUE), 2)
+
 mean_violent <- round(mean(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
 sd_violent <- round(sd(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
+median_violent <- round(median(cleaned_data$violent_crime_rate, na.rm = TRUE), 2)
 
@@
   Mean = c(mean_incarceration, mean_poverty, mean_unemployment, mean_violent),
+  Median = c(median_incarceration, median_poverty, median_unemployment, median_violent),
   SD = c(sd_incarceration, sd_poverty, sd_unemployment, sd_violent)
@@
 # Print the data frame as a table using kable
-kable(data, col.names = c("Variable", "Mean", "SD"), booktabs = TRUE)
+kable(data, col.names = c("Variable", "Mean", "Median", "SD"), booktabs = TRUE)
 ```
@@
 # Discussion {#sec-discussion}
-Make note that poorer nations could also be incapable of housing more individuals in jails (use kenya as case study. 284% jail capacity)
 
-## First discussion point {#sec-first-point}
+## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope
 
-## Second discussion point
+As the results in @sec-results have shown, nations with excessively high and excessively low incarceration rates have various qualities about them, outside the scope of our model, that may better explain how and why they may find themselves with the incarceration rates that they do have.
+In particular, various cultural differences across nations such as differences in how nations view the idea of incarceration could significantly impact how various countries choose to deal with crime.
+
+Countries such as Norway prioritize rehabilitation over punitive measures, leading to recidivism rates as low as 20%.
+On the contrary, the United States - who focus much more heavily on punishment than rehabilitation - finds over two-thirds of released prisoners returning back to the criminal justice system within 2 years [@citeNorwayUS].
+Such stark differences in how nations choose to deal with crime can vastly impact how likely they are to incarcerate an individual, as well as how likely incarcerated individuals are to return to crime.
+In essence, prioritizing punishment as a deterrent can create a vicious cycle of high incarceration rates subsequently leading to high crime rates, in turn necessitating the need for more punishment, continuing the cycle.
+
+Thus, it could appear that countries who suffer from mass incarceration, such as the United States, may find themselves in these positions by their own volition, as they choose to focus on harming their prisoners than helping.
+
+## Financial hardships and corruption at the national level may prevent nations from incarcerating when necessary
+
+The results in @tbl-incarcerationsummary indicated a negative relationship between poverty rates and incarceration rates, suggesting that nations with a generally poorer population, relative to that particular nation's cost of living, and higher income inequality may find themselves with lower incarceration rates.
+Though this may seem counter-intuitive at first glance, a potential explanation could be that poorer nations are simply less capable of handling crime.
+In particular, nations may not have the funding necessary to house criminals or hire a sufficient number of individuals to police the nation.
+For instance Kenya, who feature a poverty rate of 36.1% well above the median but an incarceration rate of 107 fairly below the median, also happen to have a jail occupancy level of 173.2% as of 2022 [@citeKenyaJail].
+This could help explain Kenya's relatively low incarceration rate as being the result of an inability to house more inmates.
+
+Another important point to raise is the effect of national poverty on the extent of corruption within a nation.
+Studies generally show that the levels of corruption in developing nations outweigh those of developed nations [@citeCorruption].
+As a result, it may be the case that poorer nations find themselves with lower incarceration rates as individuals are more easily capable of avoiding imprisonment through means such as bribery.
+In turn, extreme poverty can in turn lead to a lack of incarceration both due to financial concerns as well as higher rates of corruption.
 
@@
 # Weaknesses and next steps {#sec-limitations}
+
+(Limitation, could be better to group nations into groups where they share similar cultural values and whatnot in order to account for differences in culture.)
 

### 2024-04-18 paper/paper.qmd
Finalized paper and all references.
Completed the discussion and limitations sections. Added an appendix with model checks, and added all references as well as the abstract.

@@
     geometry: margin = 1.2in
-    abstract: "Over the past five decades, the United States has witnessed a decline in the percentage of adults residing in middle-class households, alongside escalating concerns about housing affordability.
-This study aims to assess the impact of rising costs on the living standards of middle and lower-income households, recognizing them as the most affected demographic by these economic shifts.
-Utilizing data from [...], we analyze changes in income distribution and evaluate the influence of economic policy changes, including interest rates and taxation, on the standards of living of these individuals.
-Additionally, we explore how individual-level characteristics intersect with these policies to address how best to pursue the goal of economic equity across individuals of all backgrounds.
-We find that [...]."
+    abstract: "Over the last 2 decades, the United States has found itself struggling with an issue of mass incarceration, leading the world in number of individuals incarcerated.
+This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, crime, and poverty rates.
+Utilizing data from the World Population Review, and employing a Bayesian multiple linear regression model, this paper finds evidence that poverty and unemployment negatively correlate with incarceration rates, whereas violent crime rates positively correlate.
+The paper also identifies nuances in the relationship between educational attainment and incarceration rates.
+These insights may offer valuable considerations for policy-making aimed at addressing mass incarceration in the United States."
     number-sections: true
@@
 library(kableExtra)
+library(dplyr)
+library(janitor)
 
@@
 This paper utilizes various datasets from the World Population Review, in particular it utilizes their datasets on "Incarceration Rates by Country 2024", "Poverty Rate by Country 2024", "Education Rankings by Country 2024", "Unemployment by Country 2024", and "Violent Crime Rates by Country 2024".
-By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that [...]
+By conducting statistical analyses with the aforementioned data, this paper finds evidence to suggest that poverty and unemployment negatively correlate with incarceration rates, and violent crime rates correlate positively.
+Education, on the other hand, appears to have a more nuanced relationship with incarceration rates as the relationship does not appear to be consistent across all levels of education.
 
@@
 
-All data analysis was done through R [@citeR] with the aid of the following packages:
+All data analysis was done through R [@citeR] with the aid of the following packages: `tidyverse` [@citetidyverse], `here` [@citehere], `ggplot2` [@citeggplot2], `knitr` [@citeknitr], `modelsummary` [@citemodelsummary], `rstanarm` [@citerstanarm], `kableExtra` [@citekableExtra], `dplyr` [@citedplyr], and `janitor` [@citejanitor].
 
@@
 
-## Model Coefficients
+## Model Coefficients {#sec-modelcoefficients}
 
@@
 
-## Actual vs. Predicted Incarceration Rates
+## Actual vs. Predicted Incarceration Rates {#sec-actualvspredicted}
 
@@
 
-## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope
+## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope {#sec-culture}
 
-As the results in @sec-results have shown, nations with excessively high and excessively low incarceration rates have various qualities about them, outside the scope of our model, that may better explain how and why they may find themselves with the incarceration rates that they do have.
+As the results in @sec-actualvspredicted have shown, nations with excessively high and excessively low incarceration rates have various qualities about them, outside the scope of our model, that may better explain how and why they may find themselves with the incarceration rates that they do have.
 In particular, various cultural differences across nations such as differences in how nations view the idea of incarceration could significantly impact how various countries choose to deal with crime.
@@
 
-## Financial hardships and corruption at the national level may prevent nations from incarcerating when necessary
+## Economic hardships and corruption at the national level may prevent nations from incarcerating when necessary {#sec-povertycorruption}
 
@@
 
-## Third discussion point
+## Educational attainment as a nuanced metric of incarceration rates {#sec-educationincarceration}
 
-# Weaknesses and next steps {#sec-limitations}
+As was explored in @sec-modelcoefficients, we found that there was no consistent relationship between educational ranking and incarceration rates.
+In particular, though the highest educationally ranked nations found themselves with lower incarceration rates than those with poorer education, the extent to which incarceration rates were predicted to rise with a decrease in education varied depending on how much the educational ranking would fall.
+Somewhat counter intuitively, a nation falling from the highest educational category to that of the ranks "111-125" would find themselves with the highest increase in incarceration rates compared to a further worsening of education to even the lowest category of "176-198".
 
-(Limitation, could be better to group nations into groups where they share similar cultural values and whatnot in order to account for differences in culture.)
+The possible explanations for this are twofold: Firstly, as was previously examined in @sec-culture, incarceration rates can be heavily dependent on cultural aspects that were not taken into account via the model.
+As such, certain nations with exceedingly low or exceedingly high incarceration rates may be disproportionately affecting the mean incarceration rates of their respective educational ranking categories, in turn leading to inaccurate conclusions regarding the impact of education on incarceration rates.
+To illustrate this point further, El Salvador who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin.
+
+The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as explored in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education.
+
+## Insights for addressing the mass incarceration challenge in the United States drawn from the data and results
+
+By incorporating all insights derived from @sec-actualvspredicted, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:
+
+(1) The United States should stray from punitive measures as the primary means of dealing with crime as this can perpetuate a cycle of high incarceration rates, recidivism, and crime.
+Instead, shifting towards rehabilitation-oriented policies, such as those endorsed by nations like Norway, can be more effective in reducing the need for incarceration, and reduce the likelihood of repeat-offenders.
+
+(2) The United States should dedicate more resources towards enhancing their educational prowess.
+As it stands, the United States has an education ranked between "20-34", and as a result, given the overwhelming evidence that nations with the highest educational strength also happen to have the lowest incarceration rates, an increase in investment in education can aid the United States in decreasing both their violent crime rate, as well as the need for mass incarceration.
+Though we have uncovered that education can be closely linked to other factors of a nation including economic strength and standards of living, the United States is one of the most economically powerful nations, and as a result, we may come to the justified conclusion that improvements in their education are likely to yield positive results akin to that of similar nations.
+
+# Limitations and Next Steps {#sec-limitations}
+
+## Limitations
+
+There are a number of limitations to be addressed with regards to the analysis conducted and subsequent conclusions drawn.
+A significant limitation pertaining to this study is the lack of up-to-date data across all nations.
+As was discuss is @sec-data, all raw datasets contained information on nations that may not be accurate as of 2024.
+In particular, many nations had data that was collected years, and in some cases decades, ago, in turn potentially rendering much of the analysis inaccurate as significant changes could have risen over the years.
+Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various qualities of differing nations.
+As a result, the subset of countries that were included in this study may not accurately reflect the overarching global trends as only 134 out of the 193 countries recognized by the UN [@citeUN] were taken into account during the analysis.
+Thus, the subsequent conclusions drawn may be biased as the remaining countries left in the study could be skewed towards more developed nations as these may be more likely to have data readily available.
+Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whatever information was available on that particular nation.
+As a result, nations which may have reported more crime rates than others would in turn bias the results to inaccurately portray such nations as having relatively more crime than they should when compared with nations that had missing data points for certain crime statistics.
+
+A potentially significant limitation also exists with regards to the model used.
+In particular, the model does not take into account any interaction terms between the various independent variables when it likely should have, given the complex and intricate dynamics between the many variables at play.
+Though this was done to simplify the complexity of the model to allow for easier interpretations of the results, the caveat was an overall less realistic, and possibly less accurate, model.
+Another limitation with the model lies in its omission of cultural variations and associated factors across nations.
+By not grouping countries with similar cultural values or political and judicial policies, the model fails to account for factors that could greatly influence the derived results.
+As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates, a fact not taken into account by the model.
+
+## Next Steps
+
+Future analysis in the context of this study can be better improved by incorporating more up-to-date data.
+In particular, web-scraping and compiling the raw data manually as opposed to using the raw datasets used in this study could net an overall more recent, and complete, set of data points.
+Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations, and years.
+Another improvement that could be made is a better analysis of the model itself in order to fine-tune aspects such as the locations of the various priors, as well as exploring the possibility of adding additional factors and interaction terms to more accurately model incarceration rates.
 
@@
 
-# Additional data details
-
-# Model details {#sec-model-details}
+# Model details
 
@@
 
+We employ the use a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data.
+By examining the figure, it appears as though the model does not accurately fit the data, and thus amendments likely should be made as a result [@citeAlexander].
+
 ```{r}
-#| label: fig-ppcheck-postprior
+#| label: fig-ppcheck
+#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior predictive check
 #| echo: false
@@
 #| message: false
+#| warning: false
 
-pp_check(analysis_model) + theme_classic()
-
-posterior_vs_prior(analysis_model) + theme_classic() + coord_flip()
+pp_check(analysis_model) + 
+  theme_classic() +
+  theme(legend.position = "bottom")
 ```
 
+## Comparison of the posterior and prior
+
+We compare the posterior and prior in @fig-postprior to examine how the estimates change once data is taken into account [@citeAlexander].
+
 ```{r}
-#| label: fig-rhat
+#| label: fig-postprior
+#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior-prior comparison
 #| echo: false
+#| warning: false
+#| fig-pos: H
+#| message: false
+
+posterior_vs_prior(analysis_model) + 
+  theme_classic() + 
+  theme(legend.position = "bottom") +
+  coord_flip()
+```
+
+## Markov Chain Monte Carlo Convergence
+
+As `rstanarm` makes use of the sampling algorithm MCMC [@citeAlexander], we examine whether the algorithm ran into issues through the use of an Rhat plot and trace plot, both showcased in @fig-mcmc.
+The Rhat plot does not indicate any issues, evidenced by the values all remaining close to 1 [@citeAlexander].
+Moreover, the trace plot suggests nothing out of the ordinary given that lines remain horizontal and fluctuate in an expected manner [@citeAlexander].
+
+```{r}
+#| label: fig-mcmc
+#| fig-cap: "Checking the convergence of the MCMC algorithm"
+#| fig-subcap: ["Rhat plot", "Trace plot"]
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-pos: H
+#| layout-ncol: 1
+
+plot(analysis_model, "rhat") + 
+  theme_classic()
+
+plot(analysis_model, "trace")
+
+```
+
+## Credibility Intervals
+
+Lastly, we employ a 95% credibility interval as showcased in @fig-ci, to gain a better understanding of the probability distribution of our coefficients.
+
+```{r}
+#| label: fig-ci
+#| fig-cap: "Credible intervals"
+#| echo: false
+#| message: false
+#| warning: false
 #| fig-pos: H
 
-plot(analysis_model, "rhat")
+plot(analysis_model) + 
+  theme_classic()
+
 ```
+
 \newpage

### 2024-04-19 paper/paper.qmd
Updated abstract.
Made minor adjustment to abstract

@@
     abstract: "Over the last 2 decades, the United States has found itself struggling with an issue of mass incarceration, leading the world in number of individuals incarcerated.
-This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, crime, and poverty rates.
+This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, crime, and poverty rates, and their respective impacts on incarceration.
 Utilizing data from the World Population Review, and employing a Bayesian multiple linear regression model, this paper finds evidence that poverty and unemployment negatively correlate with incarceration rates, whereas violent crime rates positively correlate.

### 2024-04-19 scripts/01-download_data.R
Removed unnecessary files and renamed others

@@ -1,19 +1,0 @@
-#### Preamble ####
-# Purpose: Details how to acquire the raw datasets used in this paper.
-# Author: Abbass Sleiman
-# Date: 15 April 2024
-# Contact: abbass.sleiman@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: N/A
-
-#### Download data ####
-# There are 5 datasets used in analysis in total, all of which are provided by
-# the World Population Review for use. Each dataset can be manually downloaded
-# directly as a csv file. Data on incarceration rates by country can be downloaded
-# at https://worldpopulationreview.com/country-rankings/incarceration-rates-by-country,
-# data on poverty rates at https://worldpopulationreview.com/country-rankings/poverty-rate-by-country,
-# education rates at https://worldpopulationreview.com/country-rankings/education-rankings-by-country,
-# unemployment rates at https://worldpopulationreview.com/country-rankings/unemployment-by-country,
-# and violent crime rates at https://worldpopulationreview.com/country-rankings/violent-crime-rates-by-country.
-
-         

### 2024-04-19 paper/paper.qmd
Minor amendments to subheaders
Added shiny app.

@@
   - Abbass Sleiman
-thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US-Middle-Class](https://github.com/AbbassSleiman/US-Middle-Class)."
+thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US-Middle-Class](https://github.com/AbbassSleiman/US-Middle-Class). Shiny webapp can be accessed at []()"
 date: today
@@
 library(janitor)
+library(shiny)
 
@@
 All data analysis was done through R [@citeR] with the aid of the following packages: `tidyverse` [@citetidyverse], `here` [@citehere], `ggplot2` [@citeggplot2], `knitr` [@citeknitr], `modelsummary` [@citemodelsummary], `rstanarm` [@citerstanarm], `kableExtra` [@citekableExtra], `dplyr` [@citedplyr], and `janitor` [@citejanitor].
+Moreover, the shiny web app associated with this paper was made with the aid of the `shiny` package [@citeshiny].
 
@@
 
-## Posterior predictive check
+## Posterior Predictive Check
 
@@
 
-## Comparison of the posterior and prior
+## Comparison of the Posterior and Prior
 

### 2024-04-19 data/analysis_data/app.R

@@ -1,0 +1,75 @@
+# Load necessary libraries
+library(shiny)
+library(ggplot2)
+library(here)
+library(tidyverse)
+
+incarceration_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
+
+incarceration_data <- incarceration_data |>
+  mutate(education_category = factor(education_category))
+
+# Define UI
+ui <- fluidPage(
+  titlePanel("Incarceration Rate Analysis"),
+  sidebarLayout(
+    sidebarPanel(
+      selectInput("metric", "Choose a metric:",
+                  choices = c("Violent Crime Rate", "Unemployment Rate", "Poverty Rate", "Educational Ranking Category")),
+      conditionalPanel(
+        condition = "input.metric == 'Educational Ranking Category'",
+        selectInput("categories", "Choose educational ranking categories to highlight:",
+                    choices = c("1-19", "20-34", "35-52", "53-71", "72-92", "93-110", 
+                                "111-125", "126-147", "148-175", "176-198"),
+                    multiple = TRUE)
+      )
+    ),
+    mainPanel(
+      plotOutput("plot")
+    )
+  )
+)
+
+# Define server logic
+server <- function(input, output) {
+  output$plot <- renderPlot({
+    # Generate plot based on user selections
+    if (input$metric == "Violent Crime Rate") {
+      # Generate plot for violent crime rate
+      ggplot(incarceration_data, aes(x = violent_crime_rate, y = incarceration_rate)) +
+        geom_point() +
+        theme_classic() +
+        labs(x = "Violent Crime Rate", y = "Incarceration Rate")
+    } else if (input$metric == "Unemployment Rate") {
+      # Generate plot for unemployment rate
+      ggplot(incarceration_data, aes(x = unemployment_rate, y = incarceration_rate)) +
+        geom_point() +
+        theme_classic() +
+        labs(x = "Unemployment Rate", y = "Incarceration Rate")
+    } else if (input$metric == "Poverty Rate") {
+      # Generate plot for poverty rate
+      ggplot(incarceration_data, aes(x = poverty_rate, y = incarceration_rate)) +
+        geom_point() +
+        theme_classic() +
+        labs(x = "Poverty Rate", y = "Incarceration Rate")
+    } else if (input$metric == "Educational Ranking Category") {
+      # Generate plot for educational ranking category
+      selected_categories <- input$categories
+      # Reorder the educational ranking categories
[…]

### 2024-04-19 data/analysis_data/app.R
Finalized shiny app and uploaded it

@@ -1,10 +1,9 @@
 # Load necessary libraries
 library(shiny)
 library(ggplot2)
-library(here)
 library(tidyverse)
 
-incarceration_data <- read_csv(here("data/analysis_data/incarceration_data.csv"))
+incarceration_data <- read_csv("incarceration_data.csv")
 
 incarceration_data <- incarceration_data |>
   mutate(education_category = factor(education_category))
@@ -39,19 +38,19 @@ server <- function(input, output) {
       ggplot(incarceration_data, aes(x = violent_crime_rate, y = incarceration_rate)) +
         geom_point() +
         theme_classic() +
-        labs(x = "Violent Crime Rate", y = "Incarceration Rate")
+        labs(x = "Violent Crime Rate (per 100k population)", y = "Incarceration Rate")
     } else if (input$metric == "Unemployment Rate") {
       # Generate plot for unemployment rate
       ggplot(incarceration_data, aes(x = unemployment_rate, y = incarceration_rate)) +
         geom_point() +
         theme_classic() +
-        labs(x = "Unemployment Rate", y = "Incarceration Rate")
+        labs(x = "Unemployment Rate (%)", y = "Incarceration Rate")
     } else if (input$metric == "Poverty Rate") {
       # Generate plot for poverty rate
       ggplot(incarceration_data, aes(x = poverty_rate, y = incarceration_rate)) +
         geom_point() +
         theme_classic() +
-        labs(x = "Poverty Rate", y = "Incarceration Rate")
+        labs(x = "Poverty Rate (%)", y = "Incarceration Rate")
     } else if (input$metric == "Educational Ranking Category") {
       # Generate plot for educational ranking category
       selected_categories <- input$categories

### 2024-04-19 paper/paper.qmd
Updated repo link.
Updated the repo link in the paper

@@
   - Abbass Sleiman
-thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US-Middle-Class](https://github.com/AbbassSleiman/US-Middle-Class). Shiny webapp can be accessed at []()"
+thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US_Incarceration](https://github.com/AbbassSleiman/US_Incarceration).
+Shiny webapp can be accessed at [https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/](https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/)"
 date: today

### 2024-04-19 paper/paper.qmd
Minor corrections

@@
 thanks: "Code and data are available at: [https://github.com/AbbassSleiman/US_Incarceration](https://github.com/AbbassSleiman/US_Incarceration).
-Shiny webapp can be accessed at [https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/](https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/)"
+Shiny web app can be accessed at [https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/](https://wklzec-abbass-sleiman.shinyapps.io/incarceration_rate/)"
 date: today

### 2024-04-19 paper/paper.qmd
Minor adjustment.
Made minor adjustment to abstract.

@@
     abstract: "Over the last 2 decades, the United States has found itself struggling with an issue of mass incarceration, leading the world in number of individuals incarcerated.
-This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, crime, and poverty rates, and their respective impacts on incarceration.
+This paper investigates the possible causes of this by conducting a comparative analysis of global trends in education, unemployment, violent crime, and poverty rates, and their respective impacts on incarceration.
 Utilizing data from the World Population Review, and employing a Bayesian multiple linear regression model, this paper finds evidence that poverty and unemployment negatively correlate with incarceration rates, whereas violent crime rates positively correlate.

### 2024-04-19 paper/paper.qmd
Made final tweaks to paper.
Made final edits and tweaks to the paper.

@@
 Instead, it is a complex issue influenced by a multitude of factors including cultural dynamics and socioeconomic conditions.
-Thus, in an attempt to unravel the intricacies of these factors, this paper seeks to conduct a comparative analysis of global trends in education, unemployment, crime, and poverty rates as independent factors influencing a nation's incarceration.
+Thus, in an attempt to unravel the intricacies of these factors, this paper seeks to conduct a comparative analysis of global trends in education, unemployment, violent crime, and poverty rates as independent factors influencing a nation's incarceration.
 However, recognizing that the total number of incarcerations is heavily dependent on population size, this paper focuses on nations' incarceration rates as the estimand, defined to be the number of incarcerated individuals per 100,000 in a given population [@citeIncarcerationRate], which serves as a better metric for measuring the extent of imprisonment within a society.
@@
 @sec-discussion discusses the real-world implications of the results uncovered in the prior sections, and finally @sec-limitations discusses the limitations and weaknesses of the analysis conducted, as well as the next steps that could be taken to improve the overall reliability of the paper.
+Additional information regarding the model details can be found in the Appendix under @sec-model-details.
 
@@
 In particular, only countries that were present in all 5 datasets are present in the cleaned dataset, bringing the total number of nations included in this study to 134.
-Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its national poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category.
+Further, the cleaned analysis dataset was filtered to only contain the following information from the 5 datasets: a country's name, its incarceration rate (per 100k population), its poverty rate, its unemployment rate, its violent crime rate, and its educational ranking category.
 Note that a nation's violent crime rate and educational ranking category were not explicit data that were provided as part of the 5 raw datasets.
@@
 As for the educational ranking category, as opposed to treating educational ranking as a continuous variable in the analysis that will follow, we instead decided to create 10 categories representing the 10 deciles of educational ranks present in the final dataset, based on the ranks of the 134 countries present, essentially splitting the countries into 10 equally sized categories by educational ranking.
-This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, as moving up or down one rank educationally may be more subjective.
+This was done in order to have a categorical variable whose coefficient will be more easily interpreted in the model present in @sec-model, as interpreting a 1 unit increase in educational ranking can be quite arbitrary, given that moving up or down one rank educationally may be more subjective.
 @tbl-cleaned-data showcases the top 5 countries by incarceration rate in the cleaned dataset.
@@
 Violent crime on the other hand, does appear to show a somewhat positive correlation with incarceration rates.
-However what is interesting to note is that the 5 highlighted nations spanning the entire range of incarceration rates do not have a significant range of violent crime rates, at least relative to the overall range of violent crime rates in the entire dataset.
+However what is interesting to note is that the 5 highlighted nations spanning the entire range of incarceration rates do not span a significant range of violent crime rates, at least relative to the overall range of violent crime rates across the entire dataset.
 The educational ranking category appears to have a negative relationship with incarceration rates, where a higher rank, i.e.
@@
 However, somewhat interesting to note is the drop in mean incarceration rates for categories past the peak in the "111-125" ranking category.
-As such, given the fairly minimal information provided by the graphical and tabular figures above regarding the relationship between the variables of interest and a nation's incarceration rate, we employ a model in @sec-model to come to a more conclusive result.
+As such, given the fairly inconclusive information provided by the graphical and tabular figures above regarding the relationship between the chosen socioeconomic factors and a nation's incarceration rate, we employ a model in @sec-model to come to a more conclusive result.
 
@@
 As the purpose of this paper is the investigation of various socioeconomic factors and their impact on a nation's incarceration rate as a means of better understanding the United States' mass incarceration problem, we will employ a model that will better enable us to infer the relationship between these various factors and incarceration rates.
-Then using this information, I will attempt to predict what the United States' incarceration rate should be given its own individual data for each of the determinants of incarceration rate, and subsequently uncover whether they are a good estimator of incarceration rate, or whether the problem stems from a different facet.
+Then, using this information, I will attempt to predict what every nation's, including the United States', incarceration rate should be given its own individual data for each of the determinants of incarceration rate, and subsequently uncover whether these socioeconomic factors are a good estimator of incarceration rate, or whether issues regarding mass incarceration stem from a different facet.
 
@@
 
-The particular model that this paper will utilize is a Bayesian multiple linear regression model of incarceration rate as a function of the independent variables present in our analysis.
+The particular model that this paper will utilize is a Bayesian multiple linear regression model of incarceration rate as a function of the socioeconomic variables present in our analysis.
 In particular, the model is as follows:
@@
 On the contrary, as predicted, violent crime rates correlate positively with incarceration rates, all else held constant.
-As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the highest category of education, "1-19", would see a much larger increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing smaller increase near the lowest ranked nations.
+As for a nation's educational ranking, it appears that there is no clear relationship between education and incarceration rates as nations in the highest category of education, "1-19", would see a much larger increase in predicted incarceration rates as they enter categories near the median educational ranking, before then seeing smaller increases near the lowest ranked nations.
 
@@
 
-## Nations with extreme incarceration rates necessitate the need to consider additional factors beyond the model's scope {#sec-culture}
+## Nations with extreme incarceration rates highlight the importance of cultural factors in determining their incarceration rates {#sec-culture}
 
@@
 Though this may seem counter-intuitive at first glance, a potential explanation could be that poorer nations are simply less capable of handling crime.
-In particular, nations may not have the funding necessary to house criminals or hire a sufficient number of individuals to police the nation.
-For instance Kenya, who feature a poverty rate of 36.1% well above the median but an incarceration rate of 107 fairly below the median, also happen to have a jail occupancy level of 173.2% as of 2022 [@citeKenyaJail].
+In particular, poorer nations may not have the funding necessary to house criminals or hire a sufficient number of individuals to police the nation.
+For instance Kenya, who feature a poverty rate of 36.1% - well above the median - but an incarceration rate of 107 - fairly below the median, also happen to have a jail occupancy level of 173.2% as of 2022 [@citeKenyaJail].
 This could help explain Kenya's relatively low incarceration rate as being the result of an inability to house more inmates.
@@
 As such, certain nations with exceedingly low or exceedingly high incarceration rates may be disproportionately affecting the mean incarceration rates of their respective educational ranking categories, in turn leading to inaccurate conclusions regarding the impact of education on incarceration rates.
-To illustrate this point further, El Salvador who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin.
+To illustrate this point further, El Salvador, who feature the highest incarceration rate of 1086, also happen to be in the educational ranking category of "111-125", which may explain why the model predicts that this particular category is likely to raise a nation's incarceration rate by the greatest margin.
 
-The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as explored in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education.
+The second possible explanation could be that nations with poorer levels of education are likely to also be poorer economically, which, as discussed in @sec-povertycorruption, could lead to lower incarceration rates, in turn providing a possible explanation for why nations with the lowest levels of education also happen to feature lower incarceration rates than some nations with better education.
 
@@
 
-By incorporating all insights derived from @sec-actualvspredicted, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:
+By incorporating all insights derived from @sec-culture, @sec-povertycorruption, and @sec-educationincarceration, we arrive to the following conclusions for how the United States can best address their issue of mass incarceration:
 
@@
 In particular, many nations had data that was collected years, and in some cases decades, ago, in turn potentially rendering much of the analysis inaccurate as significant changes could have risen over the years.
-Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various qualities of differing nations.
+Moreover, not all datasets included information on all nations, and in many cases, countries that were included in some datasets featured missing data points for various columns of data.
 As a result, the subset of countries that were included in this study may not accurately reflect the overarching global trends as only 134 out of the 193 countries recognized by the UN [@citeUN] were taken into account during the analysis.
 Thus, the subsequent conclusions drawn may be biased as the remaining countries left in the study could be skewed towards more developed nations as these may be more likely to have data readily available.
-Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whatever information was available on that particular nation.
+Moreover, due to the missing data points for countries that were included in the study, values such as the violent crime rate statistic that was calculated for each nation is inaccurate as it only included whichever information was available for that particular nation.
 As a result, nations which may have reported more crime rates than others would in turn bias the results to inaccurately portray such nations as having relatively more crime than they should when compared with nations that had missing data points for certain crime statistics.
@@
 By not grouping countries with similar cultural values or political and judicial policies, the model fails to account for factors that could greatly influence the derived results.
-As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates, a fact not taken into account by the model.
+As we have already explored in @sec-culture, varying cultural and societal norms with regards to how nations deal with crime can significantly impact their incarceration rates - a fact not taken into account by the model.
 
@@
 In particular, web-scraping and compiling the raw data manually as opposed to using the raw datasets used in this study could net an overall more recent, and complete, set of data points.
-Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations, and years.
+Moreover, incorporating time series data spanning multiple years across all nations would allow us to account for time and country fixed-effects, in turn providing more conclusive data that takes into account the various cultural and societal disparities across nations and years.
 Another improvement that could be made is a better analysis of the model itself in order to fine-tune aspects such as the locations of the various priors, as well as exploring the possibility of adding additional factors and interaction terms to more accurately model incarceration rates.
@@
 
-# Model details
+# Model details {#sec-model-details}
 
@@
 
-We employ the use a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data.
+We employ a posterior predictive check in @fig-ppcheck to evaluate how well the model fits the data.
 By examining the figure, it appears as though the model does not accurately fit the data, and thus amendments likely should be made as a result [@citeAlexander].
@@
 #| label: fig-ppcheck
-#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior predictive check
+#| fig-cap: Examining how the model fits the data through a posterior predictive check
 #| echo: false
@@
 #| label: fig-postprior
-#| fig-cap: Examining how the model fits, and is affected by, the data through a posterior-prior comparison
+#| fig-cap: Examining how the model is affected by the data through a posterior-prior comparison
 #| echo: false
